Collaborative Text Editing without Operational Transformation

### 2012-12-13 paper2009.tex
first commit

% Comments --
% 12/10/12 -
%  - incorporate previous work on CRDTs by Preguica, Weiss,  Urso, Molli, Oster, etc.
%
% 03/01/10 - 
%   -Benchmark space complexity of Node extension vs node of size one
%   -Explain the three types of linked lists (section 6)
%   -Work the clash example with the hashtable for node ids, linked list of
%    of global subnodes, linked list of local subnodes.
%   -Create git repository for Dissertation and Collabed.
%   -
% 02/23/10 - Work on the optimization section and lead it into experimental
% results.

% Future work: Intention preserving; Why our approach could really preserve
% a cut and paste (via tree-node transplant) and OPT can't

% 02/17/10 - Starting a new collabed session of the MSET paper.
%     This paper, along with all the images and scripts are
%     stored in a git repository on pythia as "paper-collabed.tex"
%     at address "pythia.cs-i.brandeis.edu/git/mset_paper"

% 7/28 - completed first draft of the paper through the algorithm
%     we still need to add more figures and to revise to make it
%     more readable

@@

\title{Optimally Efficient Collaborative Text Editing}
%
%

@@

\begin{abstract}
In this paper we present a new collaborative text editing data type,  
the Monotone Shared Edit Tree (MSET). This algorithm is an extension
and optimization of the TreeDoc datatype of Preguica and Shapiro.  The MSET data type is an example of
a Commutative Replicated Data Type. It is not based on operational transformation 
and does not require a central server to serialize the
edit operations.
MSET allows multiple users to locally 
edit their own copies of a shared text document
while simultaneously sending out their local edits 
to remote users and processing edits from remote users 
on their local copy.
Each edit operation (whether generated locally or remotely) 
requires $O(k\log(M))$ time to process, where
M is the total number of characters that have inserted 
into the user's document before the operation in
question is applied (including those characters that were later deleted.)
The MSET algorithm is based on 
the Model-View-Controller paradigm in which the
model underlying the text document is a simple and 
natural type of tree that represents the full collaborative edit process.
The edits that are allowed for this tree are monotone in the sense that
the individual edits add a component to the tree which can never be removed
though it can be evolved. For example, deletion is handled by setting the
visibility attribute of the character to ``hidden''. The operations are also
designed to facilitate shared editing in the sense that the order in which
they are applied does not affect the eventual result.
The MSET algorithm is convergent in the sense that 
if all users stop generating
editing operations, then when all operations have 
been received and processed, all users will have
the same document. 
A collaborative editor based on the MSET algorithm 
has been successfully introduced as a plugin into widely used
editors such as JEdit and Netbeans and 
these collaborative editors have been used in classroom situations
with positive feedback from the users (both students and faculty).
\end{abstract}
\newpage
\tableofcontents
\newpage

\section{Introduction}
Collaborative text editing has a long history stretching back to the dOPT model of Ellis and Gibbs from 1989 \cite{ellis_concurrency_1989}. Most of the early work on collaborative editing used an operational transformation approach with a central server that would serialize all edit operations and rebroadcast them to the clients \cite{nichols_high-latency_1995}. The server and clients would transform the edit operations before applying them to account for differences in the remote documents when the original edit operations were applied. Many extensions and variations of this early work have been developed over the ensuing 25 years (see e.g. \cite{sum_operational_2004}).

In the past few years, a new approach has been developed that greatly simplifies the process of collaborative editing by relying on data types in which the edit operations are commutative and hence do not need to be transformed. These include Uniwiki \cite{oster_building_2010, oster_uniwiki_2009},  TreeDoc \cite{ letia_consistency_2010, preguica_commutative_2009}, WOOT \cite{oster_data_2006}, LOGOOT \cite{weiss_logoot:_2008, weiss_logoot:_2009, weiss_logoot-undo:_2010}, WOOKI \cite{weiss_wooki:_2007}.

Our work, MSET, is a generalization and optimization of the TreeDoc data type  of Preguica, et. al.  \cite{preguica_commutative_2009}. It has been available since 2007 as an open source project hosted at sourceforge (http://collabed.svn.sourceforge.net/viewvc/collabed/src/) and has been used as the foundation of the CollabEd system  \cite{granville_collabed:_2009}, which provides a common plugin architecture for code editors allowing developers on different platforms (e.g. jEdit, NetBeans, Eclipse) to collaboratively edit a code file while still using the IDE that they prefer.

In the rest of this paper we give an overview of the central problems faced by collaborative editors and then we introduce the MSET data type.  The first version is an unoptimized data type which is a simple extension of the TreeDoc model. We then show how to optimize the operations so that every insert or delete operation has time complexity $O(k\log(M))$ for the client originally performing the operation and for each client that receives the broadcasted operation. In contrast to the approach used in TreeDoc, we do not need to rebalance the tree to obtain optimal performance.

@@

\section{Overview of the MSET algorithm}

@@

@@

Let $u$ be the user performing one of the following tree tree edit operations
on the target node {\tt v/m:p}.
\begin{itemize}
\item ${\tt treeextend}(v/m:p,\alpha)\;\;\text{if u=v}$ \newline
First, user $u$ can extend the text in the node $v/m$
provided it is the owner (that is $v=u$). It extends the node by appending
some text $\alpha$ to the location $p$ at the end of the node,
Note however, that users can not modify
the text in any node that they do not own and can not change any of the text
once it has been put into the node.
\item ${\tt treeinsert}(v/m:p,q,u/n,\alpha) \;\; \text{if $q\le p$}$
\newline
Second, user $u$ can insert a new node $u/n$ containing a string $\alpha$
at any position $q\le p$ of the node, 
$v/m$ provided
they have not already inserted a node there.
\item ${\tt treehide}(v/m:p,q)  \;\; \text{if $q\le p$}$\newline
Third, user $u$ can set the visibility of a character at position $q < p$
in  $v/m:p$ to false.
When characters are initially added to a node they have visibilty true, but once
set to false, the visibilty must remain false. 
\end{itemize}
For all of these operations, we say that the node $v/m:p$ is the target of the edit operation.

Given these rules we see that a group of distributed users can easily edit a 
shared tree by keeping their own representation of the tree and broadcasting
their edits to their peers. Observe that these edit operations commute in
the following sense. The only place where two user's could have a editing
conflict is when they both want to insert a string of characters at a particular
position $q$ in some node $v/m:p$, but in this case their insertions are always
ordered by userid and this resolves the conflict. No matter which order the
operations are applied the same tree results. This is phrased more formally
in the following lemma:

@@

\begin{proof}
First, the sequence $S_{i}=\sigma_i(S_{i-1})$ is well-defined because we have
stipulated that the target of each $\sigma_i$ must exist in $S_{i-1}$ so the
operator can be applied. To see that the two resulting trees are equivalent
we simply observe that the extend operations for a node $v/m:p$ must appear
in the same order in the $\sigma_i$ as the $\tau_i$ and since they effect only
the node $v/m$ the node $v/m$ will contain the same string in both $S_n$ and $T_n$.
The insert operations for a given node $v/m$ on the other hand can appear
in any order and their effect will be to add children to positions $q$
of the final node. When multiple insertions attach to the same position, they
will be ordered by user id. Since they are inserted into an ordered list,
the order in which they are inserted doesn't matter.
\end{proof}

@@

\begin{proof}
This follows easily from the previous lemma where we let $\tau_1,\ldots,\tau_m$
denote the sequence of operations performed by user $1$ and let 
$\sigma_1,\ldots,\sigma_m$ denote the operations performed by user $i$. Then
these are clearly valid rearrangements of each other and by the previous
lemma the resulting trees are the same.
\end{proof}

\section{The unoptimized model for edit trees with a edit-string view}
We are really interested in collaborative editing of strings, not trees, so the
question arises as to how we can convert string edit operations of the user
into tree edit operations on the user's model, and how we can convert tree edit
operations from the user's peers into string edit operations on the user's string,
and how can we efficiently broadcast these edit operations to all other members
of the collaborative edit session.

@@

We describe the valid edit-string insertion operations as insertions
between two symbols $d_i$ and $d_{i+1}$ in the edit-string, where the $d_i$
is one of the following:
\begin{itemize}
\item  a visible character (denoted $c_i$),
\item a hidden character (also denoted $c_i$),
\item  a start marker $<^v_m$ or
\item an end marker $>^v_m$
\end{itemize}
Let $u$ denote the user performing
the operation and let $n$ be the next available block id for user $u$, so if
the user inserts a new block containing the string $\alpha$, it will be inserted
as $<^u_n \alpha >^u_n$.  Let $c_1$ and $c_2$ denote non-marker characters
which could be either visible or hidden or one of each and let
 $v$ and $w$ be users different from $u$ and each other.

\begin{lemma} Notation as above. The following are precisely the set of valid insert operations, where we indicate the two elements on either side of the insert on the left and the result on the right, together with the condition (if any) that must be satisfied for that insertion to be allowed:
\begin{eqnarray}
c_1\;\;\;\;\; c_2 &\rightarrow& c_1 <^u_n \alpha >^u_n c_2 \label{eq:ins1}\\
c_1\;\; >^w_m &\rightarrow& \left\{ 
  \begin{array}{l l}
      c_1 \alpha >^w_m &  {\rm if}\; u=w  \\
      c_1 <^u_n \alpha >^u_n >^w_m &  {\rm if} \; u \ne w 
  \end{array}  \right . \label{eq:ins2}\\
c_1 \;\;<^w_m &\rightarrow& \left\{ 
\begin{array}{l l}
c_1 <^u_n \alpha >^u_n <^w_m & {\rm if}\; u<w  \\
FAIL & {\rm if} u \ge w \\
\end{array}
\right . \label{eq:ins3}\\
>^v_r \;\;c_2 &\rightarrow& 
 \left\{ 
  \begin{array}{l l}
    >^v_r  <^u_n \alpha >^u_n c_2 & {\rm if} \; v<u \\
    FAIL & if v \ge u \\
  \end{array} \right .\label{eq:ins4}\\
>^v_r \;\;>^w_m &\rightarrow& 
  \left\{
    \begin{array}{l l}
      >^v_r \alpha >^w_m &{\rm if}\; u=w \\
      >^v_r <^u_n \alpha >^u_n >^w_m & {\rm if} \; (v<u)  \&  (v \ne w) \\
     FAIL & {\rm (v > u) \& (v \ne w)}
    \end{array}\right . \label{eq:ins5} \\
>^v_r \;\;<^w_m &\rightarrow & 
\left \{
\begin{array}{l l}
 >^v_r <^u_n \alpha >^u_n <^w_m & {\rm if} \; v<u<w \\
FAIL & {\rm otherwise} 
\end{array} \right . \label{eq:ins6}\\
<^v_r \;\;c_2 &\rightarrow& <^v_r  <^u_n \alpha >^u_n c_2\label{eq:ins7}\\
<^v_r \;\;>^w_m &\rightarrow& 
\left \{
\begin{array}{l l}
<^v_r <^u_n \alpha >^u_n >^w_m & {\rm if} \; v=w=0, r=m=1\\
FAIL & {\rm otherwise} \\
\end{array} \right . \label{eq:ins8}\\
<^v_r \;\;<^w_m &\rightarrow&
\left \{
\begin{array}{l l}
 <^v_r <^u_n \alpha >^u_n <^w_m \;\;\; {\rm if} \; u<w \\
FAIL & {\rm otherwise} \\
\end{array} \right . \label{eq:ins9}
\end{eqnarray}
\end{lemma}

Note that user $u$ can always insert a new node between any two non-marker
characters (Eq. \ref{eq:ins1}), and user $u$ can always insert a string
directly into the end of a node that it owns ($u=w$ cases in Eqs \ref{eq:ins2}, \ref{eq:ins5}).
Also, the only empty node is the root of the tree $<^0_1\;\; >^0_1$ and this
is the only time you have a start marker directly followed by a end marker
(Eq. \ref{eq:ins8}).  The final case to take note of is Eq. \ref{eq:ins7} where
$u$ inserts a string at the between a start-marker and a non-marker characters.
We will come back to these five insertion cases in the next section.

For now, we observe that if a edit-string view only allows the insertions
as shown above for user $u$, then each such insertion corresponds to a valid
edit-tree operation and hence the collaborative editing protocol in the previous
section for edit-trees can be naturally extended to edit-strings. The following
lemma states the convergence property for this distributed algorithm.

@@

\section{The unoptimized model for edit trees with a standard-string view}

@@

If $\gamma$ is a edit-tree and $f(\gamma)$ denotes the standard string view
of that tree, then we want to show that for each string operation $\sigma$
on $f(\gamma)$ there is a corresponding valid operation $\tau$
on the edit-tree, such that 
\[
f(\tau(\gamma)) = \sigma(f(\gamma))
\]
This will allow us to convert string operations into edit-tree operations which
can then be broadcast to all user as in the previous section. There may be many possible $\tau$ operations that satisfy this property (and often there are many), but all we need is the existence of at least one such operation. At this point, we don't worry about the computation complexity of calculating the edit-tree operation corresponding to a string operation, that will be the subject of the next section.

@@

The insert operation is a bit trickier mainly because, as we have
seen in the example at the beginning of this paper,
a single insert position in
the standard view could correspond to many different insert positions in the 
edit-tree, not all of them legal.

@@

So far we have described a relatively simple collaborative editing algorithm and
proved that it is correct . We have not shown however how
to implement it efficiently. A naive implementation would require time $O(N)$ for
each step, where $N$ is the size of the edit-string. (This would happen
for example if the edit-tree is comb-shaped which would require time $O(N)$
to traverse from the root to the farthest leaf.)

@@

\section{Optimizing the model}
% Here we describe the subnode implementation with red-black trees and the initial and
% final subnodes for each node. We prove that the operations are optimally efficient.

We have shown that edit-trees can be used to implement a collaborative editor
which allows the user to immediately apply local edits and which has the convergence
property.  The model as presented in the previous section becomes unacceptably slow as the size of the tree grows.  In this section, we show how to optimize this model by creating a more complex representation which allows
the string to tree and tree to string conversions on edit operations to be computed
in time $O(k\log(N))$ where $N$ is the size of the edit tree and $k$ is the number
of characters in the operation (i.e. number inserted or deleted).

In particular, the MSET algorithm will maintain the edit tree as described above,
but will also maintain an auxiliary data structure $M$ that allows it to rapidly
compute the node position in the edit tree corresponding to an element (i.e. a visible or hidden character or a start or end marker) in any one
of three string views, and vice versa. The three string views are
\begin{itemize}
\item $M.S$: the list of all visible characters
\item $M.T$: the list of all visible and hidden characters
\item $M.E$: the list of all edit-string characters which faithfully represents the edit-tree by showing all visible and hidden characters and all markers (both
start markers and end markers)
\end{itemize}
We will show that these three lists can be implemented in such a way as to be able
to calculate the $k$th element of each list (e.g. M.E[k]) in time $O(\log(N))$ where $N$ is the size of the edit-tree.

We will represent each of the characters (visible, hidden, or marker) uniquely as an
element of type Atom.  With this approach, the lists $S$ and $T$ are sublists of $E$.
An Atom element will have the following fields:
\begin{itemize}
\item $nodepos$ this takes values of the form $(u/i/p)$ which specify the position $p$ of the atom in the $ith$ node created by user $u$.
\item $visible$ - this boolean value is true if the character is visible, false otherwise. It is false for the marker symbols and characters are "deleted" by setting this field to false.
\item $character$ - this is the character at that position which is either a character typed in by the user, or is a start or end marker $<^u_i$.
\item $node$ - this is a pointer to the particular node in the edit tree containing the character.
\end{itemize}

Each node $N$ in the edit tree will be represented by the following components:
\begin{itemize}
\item $N.user$
\item $N.nodecount$
\item $N.atom$ which is an arraylist where $N.atom[i]$ will be a link to the $i$th atom in the node, where $i$ is a position consisting either of a non-negative integer or one of the values $startmarker$ or $endmarker$.
\item $N.insertionSet$ where $N.insertionSet[k]$ is a balanced binary tree map $S$ of the nodes inserted at position $k$ in the node, and that $S(u)$ for a user $u$ can be calcuated in time $O(\log(m))$ where $m$ is the number of nodes inserted at position $k$, and that $S(u)$ gives the node owned by user $u$ and inserted at that position, if it exists; or gives a pair of nodes $n1,n2$ on either side of the position where a node owned by user $u$ should be inserted. We need this data structure to be able to rapidly process remote insertions with conflicts at the same position in a node. 
\end{itemize}
Each insertion set can be implemented as a treemap from userids to nodes.

We assume that $N$ can be used to find the
Likewise, we must be able to rapidly compute the inverse functions $g_s(n,o)$, $g_t(n,o), g_e(n,o)$ which give the offset in the specified string ($s,e,n$) corresponding to the specified node $n$ and offset $o$ within that node.

The position of a character in the edit-tree is specified by the node ($n(u,i)$ where $u$ is the unique user id and $i$ is a counter unique for that user) and by the offset $j$ within that node. Recall that for the $i$th node of user $u$,  offset 0 corresponds to the start marker $<^u_i$ and that last position $k+1$ in the node corresponds to the end marker $>^u_i$, assuming there are $k$ characters (visible and hidden) in that node.

%To maintain $M$ one needs to
%explicitly specify information about which node positions are entered in the full
%tree.  We will create node position objects which contain information about 
%which node they are in and in which position (e.g. either an index or, for
%markers, the start or end position). Observe that a {\tt nodepos} is specified
%by $u/n-p$ where $u/n$ is a node and $p$ is the position of a character
%in that node. To be precise, we let $p=0$ denote position of the start marker,
%and $p=i$ for the $i$-th character in the node, and $p=K+1$ for the end merker
%where there are $K$ non-marker characters in the node. The data structure $M$ is
%represented as a doubly linked list of the node positions as they would appear 
%in the full-edit-tree.

There are two lookup operations:
\begin{itemize}
\item {\tt M.lookupNodePos(strpos,view)} - return the node position corresponding
to the character at the specified string position in the specified view.

\item {\tt M.lookupStrPos(nodepos,view)} - return the string position in the
specified view of the character at the specified nodepos in the edit-tree.
\end{itemize}
where {\tt view} specifies the current view and is either {\tt standard} ($s$),
{\tt tracking} ($t$) or {\tt edit-tree} ($e$) as described above.

There are also two insertion operations which must be called when
new characters and/or nodes are added to the edit-tree. 
These add the newnodepos, directly before or after the character at the specified nodepos in the full-edit-tree view of the edit-tree.
\begin{itemize}

@@

\end{itemize}
which must be called whenever we hide a node $(n(u,i),p)$ in the edit-tree.

\subsection{Efficient implementation of $M$}
The data type $M$ can be implemented for each of the three views separately
by maintaining a balanced binary tree whose leaves are the node positions
in the current edit tree and where each internal node stores the number
of characters in the subtree emanating from that node. One can find the
node position corresponding to a string offset by traversing from the root
down to a leaf. Likewise, one can find the offset of a node position by traversing
upward from the corresponding leaf to the root. Hiding a node can be performed
by traversing from the corresponding leaf to the root and subtracting one
from each intermediate node for the standard view. Inserting nodes before
or after a given node requires adding a node and rebalancing. The algorithm
now depends on which kind of balanced trees are used.

%%% This needs more explanation
%%%
Observe that $M$ can be viewed as maintaining three doubly linked lists
where it takes time $O(\log(N))$ to traverse from one list element to the
previous or next.

@@

We also need some data structures to represent the nodes efficiently.
Each node is represented by 
\begin{itemize}
\item an arraylist of attributed characters (visible or hidden or marker)
\item an arraylist of {\bf insertion sets} for each position in the node. 
Nothing can be inserted at position 0 as that is before the start marker.
\end{itemize}
An arraylist is a data structure that allows one array access to its elements
and which automatically grows the array when needed.

Each {\bf insertion set} allows one to quickly  insert a new node 
in the insertion set (i.e. in time $O(\log(N))$)
and to quickly (i.e. in time $O(\log(N))$ find the node (if any) that appears before and/or
after the newly inserted node. We can use a binary tree map indexed by
the userid to implement an insertion set.

When processing remote edit operations, we will need to rapidly lookup
the node of the edit tree corresponding to a given $u/n$. For this purpose
we can use a balanced binary tree using the lexicographical ordering on
node ids $u/n$. This takes time $O(\log(U*N)$ to lookup the node from its
name, where $U$ is the number of users and $N$ is the total number of insert operations
applied by all of the users.  
In practice, for a bounded number of users, 
we could also use a hash table and get time $O(1)$

@@

If $u/n$ is not in the hashtable, then we must add $\tau$ to a queue of
operations that depend on $u/n$.
Each time a new node is added to the hashtable, we can check to see if any
operations are waiting on that node. If so, then we add that list of operations
to the incoming queue (preferably at the front of the queue!)

@@

@@

%%%%  Work getting parts of the rest of this into the main article
%\newpage
%\section{Details of a Pratical M}
%In practice, we don't use node positions of size 1 but rather use subnodes
%of size 1 or more that consist of sequences of characters that all share the
%same attribute (e.g. hidden or visible). This complicates the algorithms
%as deletion requires one to split subnodes which are then reinserted into
%the data structure $M$, but it greatly decreases the speed of the algorithm
%(not asymptotically of course). Moreover, we don't introduce explicit
%start or end marker nodes of size zero unless they are needed, which only
%occurs if there is an insertion at the beginning or at the end of a node.
%
%Also, discuss algorithms for preferring extensions rather than
%inserts where possible.
%
%%% KGG needs to change data references to characters to maintain consistency
%
%\subsection{Introducing Subnodes}
%$Subnode$: A reference to a segment of an MSET node's sequence of characters that

>>>>

% Comments --
% 7/28 - completed first draft of the paper through the algorithm
%     we still need to add more figures and to revise to make it
%     more readable

@@

\title{Collaborative Text Editing without Operational Transformation}
%
%

@@

\begin{abstract}
In this paper we present a new collaborative text editing algorithm,  
the Monotone Shared Edit Tree (MSET)
which is not based on operational transformation 
and does not require a central server to serialize the
edit operations.
The MSET algorithm allows multiple users to locally 
edit their own copies of a shared text document
while simultaneously sending out their local edits 
to remote users and processing edits from remote users 
on their local copy.
Each edit operation (whether generated locally or remotely) 
requires $O(k\log(M))$ time to process, where
M is the total number of characters that have inserted 
into the user's document before the operation in
question is applied (including those characters that were later deleted.)
The MSET algorithm is based on 
the Model-View-Controller paradigm in which the
model underlying the text document is a simple and 
natural type of tree that represents the full collaborative edit process.
The edits that are allowed for this tree are monotone in the sense that
the individual edits add a component to the tree which can never be removed
though it can be evolved. For example, deletion is handled by setting the
visibility attribute of the character to ``hidden''. The operations are also
designed to facilitate shared editing in the sense that the order in which
they are applied does not affect the eventual result.
The MSET algorithm is convergent in the sense that 
if all users stop generating
editing operations, then when all operations have 
been received and processed, all users will have
the same document. 
A collaborative editor based on the MSET algorithm 
has been successfully introduced as a plugin into widely used
editors such as JEdit and Netbeans and 
these collaborative editors have been used in classroom situations
with positive feedback from the users (both students and faculty).
\end{abstract}
\newpage
\tableofcontents
\newpage

\newpage
\section{Introduction}
Collaborative text editing has a long history stretching back to the dOPT model of ... in the early 1980s.
Many extensions and variations of this early work have been developed over the ensuing 25 years including
...

@@

\newpage
\section{Overview of the MSET algorithm}

@@

\newpage

@@

Let $u$ be the user performing one of the following tree tree edit operations
on the target node {/tt v/m:p}.
\begin{itemize}
\item ${\tt treeextend}(v/m:p,\alpha)\;\;\text{if u=v}$ \newline
First, user $u$ can extend the text in the node $v/m$
provided it is the owner (that is $v=u$). It extends the node by appending
some text $\alpha$ to the location $p$ at the end of the node,
Note however, that users can not modify
the text in any node that they do not own and can not change any of the text
once it has been put into the node.
\item ${\tt treeinsert}(v/m:p,q,u/n,\alpha) \;\; \text{if $q\le p$}$
\newline
Second, user $u$ can insert a new node $u/n$ containing a string $\alpha$
at any position $q\le p$ of the node, 
$v/m$ provided
they have not already inserted a node there.
\item ${\tt treehide}(v/m:p,q)  \;\; \text{if $q\le p$}$\newline
Third, user $u$ can set the visibility of a character at position $q < p$
in  $v/m:p$ to false.
When characters are initially added to a node they have visibilty true, but once
set to false, the visibilty must remain false. 
\end{itemize}
For all of these operations, we say that the node $v/m:p$ is the target of the edit operation.

Given these rules we see that a group of distributed users can easily edit a 
shared tree by keeping their own representation of the tree and broadcasting
their edits to their peers. Observe that these edit operations commute in
the following sense. The only place where two user's could have a editing
conflict is when they both want to insert a string of characters at a particular
position $q$ in some node $v/m:p$, but in this case their insertions are always
ordered by userid and this resolved the conflict. No matter which order the
operations are applied the same tree results. This is phrased more formally
in the following lemma:

@@

\begin{proof}
First, the sequence $S_{i}=\sigma_i(S_{i-1})$ is well-defined because we have
stipulated that the target of each $\sigma_i$ must exist is $S_{i-1}$ so the
operator can be applied. To see that the two resulting trees are equivalent
we simply observe that the extend operations for a node $v/m:p$ must appear
in the same order in the $\sigma_i$ as the $\tau_i$ and since they effect only
the node $v/m$ the node $v/m$ will contain the same string in both $S_n$ and $T_n$.
The insert operations for a give node $v/m$ on the other hand can appear
in any order and their effect will be to add children to positions $q$
of the final node. When multiple insertions attach to the same position, they
will be ordered by user id. Since they are inserted into an ordered list,
the order ih which they are inserted doesn't matter.\qed
\end{proof}

@@

\begin{proof}
This follows easily from the previous lemma where we let $\tau_1,\ldots,\tau_m$
denote the sequence of operations performed by user $1$ and let 
$\sigma_1,\ldots,\sigma_m$ denote the operations performed by user $i$. Then
these are clearly valid rearrangements of each other and by the previous
lemma the resulting trees are the same.\qed
\end{proof}

\newpage
\section{The unoptimized model for edit trees with a edit-string view}
We are really interested in collaborative editing of strings, not trees, so the
question arises as to how we can convert string edit operations of the user
into tree edit operations on the user's model, and how we can convert tree edit
operations from the user's peers into string edit operations on the user's string,
and how can we efficiently broadcast these edit operations to all other members
of the collaborative edit session.

@@

We describe the valid edit-string insertion operations as insertions
between two symbols $d_i$ and $d_{i+1}$ in the edit-string, where the $d_i$
can be visible characters, hidden characters, or marker symbols (either start
markers $<^v_m$ or end markers $>^v_m$).  Let $u$ denote the user performing
the operation and let $n$ be the next available block id for user $u$, so if
the user inserts a new block containing the string $\alpha$, it will be inserted
as $<^u_n \alpha >^u_n$.  Let $c_1$ and $c_2$ denote non-marker characters
which could be either visible or hidden or one of each.

The following are the valid insert operations where we assume $v$ and $w$
are users
\begin{eqnarray}
c_1\;\; c_2 &\rightarrow& c_1 <^u_n \alpha >^u_n c_2 \label{eq:ins1}\\
c_1\;\; >^v_m &\rightarrow& c_1 \alpha >^u_m \;\;\; {\rm if}\; u=v \label{eq:ins2}\\
c_1\;\;>^v_m &\rightarrow& c_1 <^u_n \alpha >^u_n >^w_m \label{eq:ins3}\\
c_1 \;\;<^v_m &\rightarrow& c_1 <^u_n \alpha >^u_n <^v_m \;\;\; {\rm if}\; u<v \label{eq:ins4}\\
>^v_r \;\;c_2 &\rightarrow& >^v_r  <^u_n \alpha >^u_n c_2 \;\;\; {\rm if} \; v<u \label{eq:ins5}\\
>^v_r \;\;>^w_m &\rightarrow& >^v_r \alpha >^u_m \;\;\; {\rm if}\; u=w \label{eq:ins6}\\
>^v_r \;\;>^w_m &\rightarrow& >^v_r <^u_n \alpha >^u_n >^w_m \;\;\; {\rm if} \; v<u \label{eq:ins7}\\
>^v_r \;\;<^w_m &\rightarrow& >^v_r <^u_n \alpha >^u_n <^w_m \;\;\; {\rm if} \; v<u<w \label{eq:ins8}\\
<^v_r \;\;c_2 &\rightarrow& <^v_r  <^u_n \alpha >^u_n c_2\label{eq:ins9}\\
<^v_r \;\;>^w_m &\rightarrow& <^v_r <^u_n \alpha >^u_n >^w_m \;\;\; {\rm if} \; v=w=0, r=m=1 \label{eq:ins10}\\
<^v_r \;\;<^w_m &\rightarrow& <^v_r <^u_n \alpha >^u_n <^w_m \;\;\; {\rm if} \; u<w \label{eq:ins11}
\end{eqnarray}
Note that user $u$ can always insert a new node between any two non-marker
characters (Eq. \ref{eq:ins1}), and user $u$ can always insert a string
directly into the end of a node $u/n$ that it owns (Eqs \ref{eq:ins2}, \ref{eq:ins6}).
Also, the only empty node is the root of the tree $<^0_1\;\; >^0_1$ and this
is the only time you have a start marker directly followed by a end marker
(Eq. \ref{eq:ins10}).  The final case to take note of is Eq. \ref{eq:ins9} where
$u$ inserts a string at the between a start-marker and a non-marker characters.
We will come back to these five insertion cases in the next section.

For now, we observe that if a edit-string view only allows the insertions
as show above for user $u$, then each such insertion corresponds to a valid
edit-tree operation and hence the collaborative editing protocol in the previous
section for edit-trees can be naturally extended to edit-strings. The following
lemma states the convergence property for this distributed algorithm.

@@

\newpage
\section{The unoptimized model for edit trees with a standard-string view}

@@

If $\gamma$ is a edit-tree and $f(\gamma)$ denotes the standard string view
of that tree, then we want to show that for each string operation $\sigma$
on $f(\gamma)$ there is a corresponding valid operation $\tau$
on the edit-tree, such that 
\[
f(\tau(\gamma)) = \sigma(f(\gamma))
\]
This will allow us to convert string operations into edit-tree operations which
can then be broadcast to all user as in the previous section.

@@

The insert operation is a bit trickier mainly because, as we have
seen in the example at the beginning o this paper,
a single insert position in
the standard view could correspond to many different insert positions in the 
edit-tree, not all of them legal.

@@

So far we have described a relatively simple collaborative editing algorithm and
proved that it is correct and extremely flexible. We have not shown however how
to implement it efficiently. A naive implementation would require time $O(N)$ for
each step, where $N$ is the size of the edit-string. (This would happen
for example if the edit-tree is comb-shaped which would require time $O(N)$
to traverse from the root to the farthest leaf.)

@@

\newpage
\section{Optimizing the model}
% Here we describe the subnode implementation with red-black trees and the initial and
% final subnodes for each node. We prove that the operations are optimally efficient.

We have shown that the MSET approach can be used to implement a collaborative editor
which allows user to immediately apply local edits and which has the convergence
property.  We will speak to the "ease of use" and "intuitiveness" of the MSET algorithm
in a later section, but we still need to demonstrate that the conversions between
string operations and tree operations (and vice versa) can be performed efficiently.
The model as presented in the previous section becomes unacceptably slow as the
size of the tree grows.

We will optimize this model by creating a more complex representation which allows
the string to tree and tree to string conversions on edit operations to be computed
in time $O(k\log(N))$ where $N$ is the size of the edit tree and $k$ is the number
of characters in the operation (i.e. number inserted or deleted).

In particular, the MSET algorithm will maintain the edit tree as described above,
but will also maintain an auxiliary data structure $M$ that allows one to rapidly
compute the node position in the edit tree corresponding to a position in any one
of three string views, and vice versa. The three string views are
\begin{itemize}
\item the standard view of all visible characters
\item the full view of all visible and hidden characters
\item the edit-string view which faithfully represents the edit-tree by showing
all visible and hidden characters and all markers (both
start markers and end markers)
\end{itemize}

To maintain $M$ one needs to
explicitly specify information about which node positions are entered in the full
tree.  We will create node position objects which contain information about 
which node they are in and in which position (e.g. either an index or, for
markers, the start or end position). Observe that a {\tt nodepos} is specified
by $u/n-p$ where $u/n$ is a node and $p$ is the position of a character
in that node. To be precise, we let $p=0$ denote the first non-marker
character in the node and we use $p=start$ or $p=end$ to denote the start
and end markers (e.g. let $start=-1$ and $end=-2$)

$M$ represents a doubly linked list of the node positions as they would appear 
in the full-edit-tree.

Here are the two lookup operations:
\begin{itemize}
\item {\tt M.lookupNodePos(strpos,view)} - return the node position corresponding
to the character at the specified string position in the specified view.

\item {\tt M.lookupStrPos(nodepos,view)} - return the string position in the
specified view of the character at the specified nodepos in the edit-tree.
\end{itemize}
where {\tt view} specifies the current view and is either {\tt standard},
{\tt full} or {\tt edit-tree} as described above.

There are also two insertion operations which must be called when
new characters and/or node are added to the edit-tree. 
These add the newnodepos, directly before or after the specified nodepos in the
full-edit-tree view of the edit-tree.
\begin{itemize}

@@

\end{itemize}
which must be called whenever we hide a node in the edit-tree.

\subsection{Efficient implementation of $M$}
Observe that $M$ can be implemented for each of the three views separately
by maintaining a balanced binary tree whose leaves are the node positions
in the current edit tree and where each internal node stores the number
of characters in the subtree emanating from that node. Once can find the
node position corresponding to an string offset by traversing from the root
down to a leaf. Likewise, one can find the offset of a node position by traversing
upward from the corresponding leaf to the root. Hiding a node can be performed
by traversing from the corresponding leaf to the root and subtracting one
from each intermediate node for the standard view. Inserting nodes before
or after a given node requires adding a node and rebalancing. The algorithm
now depends on which kind of balanced trees are used. The appendix contains
a specification for the algorithm using red-black trees.

Observe that $M$ can be viewed as maintaining three doubly linked lists
where it takes time $O(\log(N))$ to traverse from one list element to the
previous or next.

@@

We also need some data structures to represent the nodes efficiently.
Each node is represented by 
\begin{itemize}
\item an arraylist of attributed characters (either visible or hidden)
\item the start and end markers for that node
\item an arraylist of {\bf insertion sets} for each position in the node. These correspond
to instances where users inserted new nodes between two nodepos elements (or at
the beginning or end of the node).
\end{itemize}
The arraylist is a datastructure that allows one array access to its elements
and which automatically grows the array when needed.

Each {\bf insertion set} allows one to quickly $O(\log(N))$ insert a new node 
in the insertion set
and to quickly $O(\log(N))$ find the nodes (if any) that appear before and/or
after the newly inserted node. We can use a binary tree map indexed by
the userid to represent an insertion set.

When processing remote edit operations, we will need to rapidly lookup
the node of the edit tree correspond to a given is $u/n$. For this purpose
we can use a balanced binary tree using the lexicographical ordering on
node ids $u/n$. This takes time $O(\log(N)$ to lookup the node from its
name.  In practice, we could also use a hash table and get time $O(1)$

@@

If $u/n$ is not in the hashtable, then we must add $\tau$ to a queue of
operations that depend on $u/n$.
Each time a new node is added to the hashtable, we can check to see if any
operations are waiting on that node. If so, then we add that list of operations
to the incomining queue (preferably at the front of the queue!)

@@

ADD FIGURES

@@

\newpage
\section{Optimizations}
In practice, we don't use node positions of size 1 but rather use subnodes
of size 1 or more that consist of sequences of characters that all share the
same attribute (e.g. hidden or visible). This complicates the algorithms
as deletion requires one to split subnodes which are then reinserted into
the data structure $M$, but it greatly decreases the speed of the algorithm
(not asymptotically of course). Moreover, we don't introduce explicit
start or end marker nodes of size zero unless they are needed, which only
occurs if there is an insertion at the beginning or at the end of a node.

Also, discuss algorithms for preferring extensions rather than
inserts where possible.

\newpage
\section{Extensions}
\subsection{Allowing more general attributes}
One simple extension to this model allows it to support general text attributes
(e.g. font-style, text-color, etc.) which can be changed multiple times by
the users while preserving the convergence property.
